Isomorphic Interpreters in {{langRevT}} (Journal Version)

### 2012-02-22 iso-int/iso-int.tex
pi diagrams
updated pi diagram
PiEE interpreter
operational semantics
_
fiore

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
%! [^ = \ulcorner
%! ^] = \urcorner
%! [v = \llcorner
%! v] = \lrcorner
%! |-->* = \mapsto^{*}
%! |-->> = \mapsto_{\ggg}
%! |-->let = \mapsto_{let}
%! |--> = \mapsto
%! |- = \vdash
%! in = \!\!\in\!\!
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! ::= = ::=
%! /= = \neq
%! vi = v_i
%! di = d_i
%! si = s_i
%! sj = s_j
%! F = \texttt{F}
%! T = \texttt{T}
%! forall = \forall
%! exists = \exists
%! empty = \emptyset
%! eta = \eta
%! where = \textbf{where}
%! epsilon = \varepsilon
%! least = \phi
%! trace+ = trace
%! trace* = trace_{\times}
%! loop+ = loop_{+}
%! loop* = loop_{\times}
%! CatC = {\mathcal C}
%! CatA = {\mathcal A}
%! gamma = \gamma
%! {[ = \{
%! ]} = \}
%! elem = \in
%! dagger = ^\dagger
%! alpha = \alpha
%! beta = \beta
%! rho = \rho
%! @@ = \mu
%! @ = \,@\,
%! langRev = \Pi
%! langRevT = \Pi^{o}
%! bullet = \bullet
%! * = \times

@@

\begin{document}
\title{On the Construction of Isomorphic Interpreters} 
\titlerunning{On the construction of Isomorphic Interpreters} 
\author{Roshan P. James \and Amr Sabry}
\institute{School of Informatics and Computing, Indiana University\\
\mails}
\maketitle

@@

%subcode{opsem} include main
% <iso, v, cc> &|-->& {[iso, iso(v), cc]}
% <c1(;)c2, v, cc> &|-->& <c1, v, Fst cc c2>
% {[c1, v, Fst cc c2]} &|-->& <c2, v, Snd c1 cc>
% {[c2, v, Snd c1 cc]} &|-->& {[ c1(;)c2, v, cc ]}
% <c1(+)c2, L v, cc> &|-->& <c1, v, LeftPlus cc c2>
% {[ c1, v, LeftPlus cc c2 ]} &|-->& {[c1 (+) c2, L v, cc ]}
% <c1(+)c2, R v, cc> &|-->& <c2, v, RightPlus c1 cc>
% {[ c2, v, RightPlus c1 cc ]} &|-->& {[c1 (+) c2, R v, cc ]}
% <c1(*)c2, (v1, v2), cc> &|-->& <c1, v1, LeftTimes cc c2 v2>
% {[ c1, v1, LeftTimes cc c2 v2 ]} &|-->& <c2, v2, RightTimes c1 v1 cc>
% {[ c2, v2, RightTimes c1 v1 cc ]} &|-->& {[ c1 (*) c2, (v1, v2), cc ]}

@@

>>>>

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
%! [^ = \ulcorner
%! ^] = \urcorner
%! [v = \llcorner
%! v] = \lrcorner
%! |-->* = \mapsto^{*}
%! |-->> = \mapsto_{\ggg}
%! |-->let = \mapsto_{let}
%! |--> = \mapsto
%! <--| = \mapsfrom
%! |- = \vdash
%! in = \!\!\in\!\!
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! ::= = ::=
%! /= = \neq
%! vi = v_i
%! di = d_i
%! si = s_i
%! sj = s_j
%! F = \texttt{F}
%! T = \texttt{T}
%! forall = \forall
%! exists = \exists
%! empty = \emptyset
%! eta = \eta
%! where = \textbf{where}
%! epsilon = \varepsilon
%! least = \phi
%! trace+ = trace
%! trace* = trace_{\times}
%! loop+ = loop_{+}
%! loop* = loop_{\times}
%! CatC = {\mathcal C}
%! CatA = {\mathcal A}
%! gamma = \gamma
%! {[ = \{
%! ]} = \}
%! elem = \in
%! dagger = ^\dagger
%! alpha = \alpha
%! beta = \beta
%! rho = \rho
%! @@ = \mu
%! @ = \,@\,
%! langRev = \Pi
%! langRevT = \Pi^{o}
%! langRevEE = \Pi^{\eta\epsilon}_{+}
%! bullet = \bullet
%! * = \times

@@

\begin{document}
\title{Isomorphic Interpreters in {{langRevT}} (Journal Version)} 
\titlerunning{On the construction of Isomorphic Interpreters} 
\author{Roshan P. James \and Amr Sabry}
\institute{School of Informatics and Computing, Indiana University\\
\mails}
\maketitle

@@

%subcode{opsem} include main
%! columnStyle = rclr
% <iso, v, cc> &|-->& {[iso, iso(v), cc]} &~~~~~~~~~~~~ rule 1
% <c1(;)c2, v, cc> &|-->& <c1, v, Fst cc c2> &~~~~~~~~~~~~ rule 2
% {[c1, v, Fst cc c2]} &|-->& <c2, v, Snd c1 cc> &~~~~~~~~~~~~ rule 3
% {[c2, v, Snd c1 cc]} &|-->& {[ c1(;)c2, v, cc ]} &~~~~~~~~~~~~ rule 4
% <c1(+)c2, L v, cc> &|-->& <c1, v, LeftPlus cc c2> &~~~~~~~~~~~~ rule 5
% {[ c1, v, LeftPlus cc c2 ]} &|-->& {[c1 (+) c2, L v, cc ]} &~~~~~~~~~~~~ rule 6
% <c1(+)c2, R v, cc> &|-->& <c2, v, RightPlus c1 cc> &~~~~~~~~~~~~ rule 7
% {[ c2, v, RightPlus c1 cc ]} &|-->& {[c1 (+) c2, R v, cc ]} &~~~~~~~~~~~~ rule 8
% <c1(*)c2, (v1, v2), cc> &|-->& <c1, v1, LeftTimes cc c2 v2> &~~~~~~~~~~~~ rule 9
% {[ c1, v1, LeftTimes cc c2 v2 ]} &|-->& <c2, v2, RightTimes c1 v1 cc> &~~~~~~~~~~~~ rule 10
% {[ c2, v2, RightTimes c1 v1 cc ]} &|-->& {[ c1 (*) c2, (v1, v2), cc ]} &~~~~~~~~~~~~ rule 11

\begin{center}
\scalebox{0.9}{
  \includegraphics{diagrams/pi1.pdf}
}
\end{center}

\begin{center}
\scalebox{0.9}{
  \includegraphics{diagrams/pi2.pdf}
}
\end{center}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{A {{langRevT}} Interpreter}

%subcode{bnf} include main
% Combinators, c = iso | c (;) c | c (*) c | c (+) c | trace c
% Combinator Contexts, cc = [] | Fst cc c | Snd c cc 
%                  &|& LeftTimes cc c v | RightTimes c v cc 
%                  &|& LeftPlus cc c | RightPlus c cc | Trace cc
% Values, v = () | (v, v) | L v | R v
%
% Machine states = <c, v, cc> | {[c, v, cc]}
% Start state = <c, v, []> 
% Stop State = {[c, v, []]}

%subcode{opsem} include main
%! columnStyle = rclr
% <iso, v, cc> &|-->& {[iso, iso(v), cc]} &~~~~~~~~~~~~ rule 1
% <c1(;)c2, v, cc> &|-->& <c1, v, Fst cc c2> &~~~~~~~~~~~~ rule 2
% {[c1, v, Fst cc c2]} &|-->& <c2, v, Snd c1 cc> &~~~~~~~~~~~~ rule 3
% {[c2, v, Snd c1 cc]} &|-->& {[ c1(;)c2, v, cc ]} &~~~~~~~~~~~~ rule 4
% <c1(+)c2, L v, cc> &|-->& <c1, v, LeftPlus cc c2> &~~~~~~~~~~~~ rule 5
% {[ c1, v, LeftPlus cc c2 ]} &|-->& {[c1 (+) c2, L v, cc ]} &~~~~~~~~~~~~ rule 6
% <c1(+)c2, R v, cc> &|-->& <c2, v, RightPlus c1 cc> &~~~~~~~~~~~~ rule 7
% {[ c2, v, RightPlus c1 cc ]} &|-->& {[c1 (+) c2, R v, cc ]} &~~~~~~~~~~~~ rule 8
% <c1(*)c2, (v1, v2), cc> &|-->& <c1, v1, LeftTimes cc c2 v2> &~~~~~~~~~~~~ rule 9
% {[ c1, v1, LeftTimes cc c2 v2 ]} &|-->& <c2, v2, RightTimes c1 v1 cc> &~~~~~~~~~~~~ rule 10
% {[ c2, v2, RightTimes c1 v1 cc ]} &|-->& {[ c1 (*) c2, (v1, v2), cc ]} &~~~~~~~~~~~~ rule 11
% < trace c, v, cc> &|-->& <c, R v, Trace cc> &~~~~~~~~~~~~ rule 12
% {[c, L v, Trace cc]} &|-->& <c, L v, Trace cc> &~~~~~~~~~~~~ rule 13
% {[c, R v, Trace cc]} &|-->& {[trace c, R v, cc]} &~~~~~~~~~~~~ rule 14

\begin{center}
\scalebox{0.9}{
  \includegraphics{diagrams/pi0-1.pdf}
}
\end{center}

\begin{center}
\scalebox{0.9}{
  \includegraphics{diagrams/pi0-2.pdf}
}
\end{center}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{A {{langRevEE}} interpreter}

%subcode{bnf} include main
% Combinators, c = iso | eta | eps | c (;) c | c (*) c | c (+) c
% Combinator Contexts, cc = [] | Fst cc c | Snd c cc 
%                  &|& LeftTimes cc c v | RightTimes c v cc 
%                  &|& LeftPlus cc c | RightPlus c cc 
% Values, v = () | (v, v) | L v | R v
%
% Machine states = <c, v, cc> | {[c, v, cc]} | <c, v, cc>^{dagger} | {[c, v, cc]}^{dagger}
% Start state = <c, v, []> 
% Stop State = {[c, v, []]}
%

%subcode{opsem} include main
%! columnStyle = rclr
% <iso, v, cc> &|-->& {[iso, iso(v), cc]} &~~~~~~~~~~~~ rule 1
% <c1(;)c2, v, cc> &|-->& <c1, v, Fst cc c2> &~~~~~~~~~~~~ rule 2
% {[c1, v, Fst cc c2]} &|-->& <c2, v, Snd c1 cc> &~~~~~~~~~~~~ rule 3
% {[c2, v, Snd c1 cc]} &|-->& {[ c1(;)c2, v, cc ]} &~~~~~~~~~~~~ rule 4
% <c1(+)c2, L v, cc> &|-->& <c1, v, LeftPlus cc c2> &~~~~~~~~~~~~ rule 5
% {[ c1, v, LeftPlus cc c2 ]} &|-->& {[c1 (+) c2, L v, cc ]} &~~~~~~~~~~~~ rule 6
% <c1(+)c2, R v, cc> &|-->& <c2, v, RightPlus c1 cc> &~~~~~~~~~~~~ rule 7
% {[ c2, v, RightPlus c1 cc ]} &|-->& {[c1 (+) c2, R v, cc ]} &~~~~~~~~~~~~ rule 8
% <c1(*)c2, (v1, v2), cc> &|-->& <c1, v1, LeftTimes cc c2 v2> &~~~~~~~~~~~~ rule 9
% {[ c1, v1, LeftTimes cc c2 v2 ]} &|-->& <c2, v2, RightTimes c1 v1 cc> &~~~~~~~~~~~~ rule 10
% {[ c2, v2, RightTimes c1 v1 cc ]} &|-->& {[ c1 (*) c2, (v1, v2), cc ]} &~~~~~~~~~~~~ rule 11
% <eps, R v, cc> &|-->& <eps, L (-v), cc>^{dagger} &~~~~~~~~~~~~ rule 12
% <eps, L (-v), cc> &|-->& <eps, R v, cc>^{dagger} &~~~~~~~~~~~~ rule 13

%subcode{opsem} include main
%! columnStyle = rclr
% <iso, v, cc>^{dagger} &<--|& {[iso, iso(v), cc]}^{dagger} &~~~~~~~~~~~~ rule 1'
% <c1(;)c2, v, cc>^{dagger} &<--|& <c1, v, Fst cc c2>^{dagger} &~~~~~~~~~~~~ rule 2'
% {[c1, v, Fst cc c2]}^{dagger} &<--|& <c2, v, Snd c1 cc>^{dagger} &~~~~~~~~~~~~ rule 3'
% {[c2, v, Snd c1 cc]}^{dagger} &<--|& {[ c1(;)c2, v, cc ]}^{dagger} &~~~~~~~~~~~~ rule 4'
% <c1(+)c2, L v, cc>^{dagger} &<--|& <c1, v, LeftPlus cc c2>^{dagger} &~~~~~~~~~~~~ rule 5'
% {[ c1, v, LeftPlus cc c2 ]}^{dagger} &<--|& {[c1 (+) c2, L v, cc ]}^{dagger} &~~~~~~~~~~~~ rule 6'
% <c1(+)c2, R v, cc>^{dagger} &<--|& <c2, v, RightPlus c1 cc>^{dagger} &~~~~~~~~~~~~ rule 7'
% {[ c2, v, RightPlus c1 cc ]}^{dagger} &<--|& {[c1 (+) c2, R v, cc ]}^{dagger} &~~~~~~~~~~~~ rule 8'
% <c1(*)c2, (v1, v2), cc>^{dagger} &<--|& <c1, v1, LeftTimes cc c2 v2>^{dagger} &~~~~~~~~~~~~ rule 9'
% {[ c1, v1, LeftTimes cc c2 v2 ]}^{dagger} &<--|& <c2, v2, RightTimes c1 v1 cc>^{dagger} &~~~~~~~~~~~~ rule 10'
% {[ c2, v2, RightTimes c1 v1 cc ]}^{dagger} &<--|& {[ c1 (*) c2, (v1, v2), cc ]}^{dagger} &~~~~~~~~~~~~ rule 11'
% {[eps, L (-v), cc]}  &<--|& {[eps, R v, cc]}^{dagger}  &~~~~~~~~~~~~ rule 12'
% {[eps, R v, cc]} &<--|&  {[eps, L (-v), cc]}^{dagger} &~~~~~~~~~~~~ rule 13'

\begin{center}
\scalebox{0.9}{
  \includegraphics{diagrams/piee1.pdf}
}
\end{center}

Let us summarize it as follows.

\begin{center}
\scalebox{1.0}{
  \includegraphics{diagrams/piee2.pdf}
}
\end{center}

Adding backward execution using the adjoint.

\begin{center}
\scalebox{1.0}{
  \includegraphics{diagrams/piee3.pdf}
}
\end{center}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{CEK machine}

%% %subcode{bnf} include main
%% % e = x | \x.e | e e
%% % rho = [] | x=v:rho
%% % E = [] | (rho, e):E | v:E
%% % v = (\x.e, rho)
%% %
%% % Machine states = <e, rho, E>
%% % Start state = <e, [], []>
%% % Stop state = <v, rho, []>

%% %subcode{opsem} include main
%% % <e1 e2, rho, E > &|-->& <e1, rho, (rho, e2):E >
%% % <\x.e, rho, (rho', e'):E > &|-->& <e', rho', (\x.e,rho):E >
%% % <\x.e, rho, (\x'.e',rho'):E > &|-->& <e', x' =(\x.e,rho):rho', E >

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\section{Isomorphic Interpreters from Algebraic manipulation}

aka. ultra-constructivism.

\begin{center}
\scalebox{1.0}{
  \includegraphics{diagrams/fiore_t3.pdf}
}
\end{center}

The following diagram expresses the isomorphism {{t^3 <-> -1}} from
Fiore's paper \cite{Fiore:2004}. The computation corresponds to the
following algebraic manipulation:

{{t^3 = t*(t*t)}}

{{t^3 = t*(t-1)}} (because {{t = t*t + 1}}  and hence {{t*t = t - 1}}).

{{t^3 = t*t-t}}

{{t^3 = -1}} (because {{t = t*t + 1}}  and hence {{t*t - t = -1}}).

However this computation is always non-terminating. So it is
computationally useless. Fiore on the other hand says that {{t^7 = t}}
has a useful computational interpretation. Which gives rise to the
conclusion that there are algebraic proofs of equalities that have
interesting computational content and once that are nonsense. What
does this mean? If these programs are proof, why are there two sorts
of proofs here? Are one set of proofs more ``constructive'' than the
other? If so, how do we characterize them?

%% data Tree = Node Tree Tree 
%%           | Leaf

%% t = t * t + 1
